Stack Safety is a Security Property

### 2021-04-15 tex/main.tex
work on 6

In the present work, we also distinguish between protecting stack data (like
LSE) and
control flow (like WBCF).  But our formulations of both properties are
somewhat different from those of~\cite{Skorstengaard+19}.  Rather than
defining stack safety implicitly,
as ``what a certain idealized machine does,'' we formalize explicit criteria
based on the technical framework of language-based
security\ifaftersubmission~\citep{??}\bcp{Maybe the Sabelfeld and Myers
  survey?}\fi.
Our core properties are \emph{stepwise stack integrity} and
\emph{stepwise stack confidentiality}.  WBCF is a desireable property, but
we can express safety in an information-flow sense without reference to it,
even safety from attacks that hijack control flow.  \BCP{On consideration
  (especially in light of the weird ``C confuses A by returning directly to
  A instead of returning to B so that B can finish up some stuff before
  returning to A'' example\footnote{Actually, this example is not so weird.
  If C, the attacker, can return a {\em result} directly to A, who is
  expecting a result from B, then that is clearly bad!}), I
agree now that we should discuss WBCF.  It should be kept as distinct as
possible from the ``data safety'' part, since it is not really an
information flow property.  But our treatment of it is interestingly
different from StkTokens, because of our call annotations, and we should
discuss that.}

@@

We are finally ready for our definition of stack safety. We begin
by describing how we model knowledge about the program structure,
then dive into the formal definitions of local state encapsulation and
well-bracketed control flow, which together constitute stack safety\bcp{not
  any more?}.
Initially we will focus on a simple stack model, with a single stack
that doesn't share between caller and callee. All arguments are passed
in registers. Recall that we keep track of which addresses are
annotated as calls, with a ``code map'':
\[\codemap \in \CODEMAPS \subseteq \WORDS\]

We parameterize over the structure of the stack, particularly the means by
which a caller identifies the data that should be protected during a call. We term
this the {\em sealing convention}, \(\sealcon \subseteq \MACHS \times \COMPONENTS\), which
maps a machine state to a set of addresses.
If a state's \(\PCname\) is annotated as a call, the sealing convention tells us
which addresses must be protected until the return. Similarly, we abstract over what
it means to return with a {\em return convention}, \(\retcon \subseteq \MACHS \times \MACHS\),
which relates call states to the states that can be considered valid returns to their caller.
The usual return convention is that \(\mach\) and \(\mach'\) agree on the stack pointer,
and \(\mach'\) has its program counter advanced by one instruction from \(\mach\).
But whatever the return convention, at each step we update the context according
to the context-step function, \(\stepsto_C\).

Each system component will be assigned to a {\em domain}, which may be {\em Outside}
of the stack entirely, {\em Unsealed} and therefore accessible, or {\em Sealed}
at a particular depth. Each caller will register a {\em target}
condition -- a predicate on machine states that must hold for it to consider itself active.
A context is a pair of a ``domain map'' from components to domains, and a
``return target list,'' which is a stack of targets, one for each caller awaiting
return.  \ifaftersubmission\bcp{spacing is funny}\fi

\[\stackDom \in \STKDOMS ::= \outside + \unsealed + \sealed{\depth}\]
\[\context \in \CONTEXTS ::= (\domMap \in \COMPONENTS \rightarrow \STKDOMS)
  \times (\rts \in \listT (\target \subseteq \MACHS)) \]

An initial context \(\context_0 = \domMap_0, []\) where \(\domMap_0\) maps
the stack to \(\unsealed\) and all other addresses and all registers to \(\outside\).

@@

When we build machine traces with this step function, we always know how deep in the
call stack we are and which components are sealed. We may now proceed to state our properties.
Our first property will be {\em stepwise stack integrity}, stepwise meaning that it will be
quantified over each step of execution from any initial state. The criterion for integrity
is simple: sealed components do not change.

\definition A system enjoys {\em stepwise stack integrity} if for any reachable state
\(\mpcstatename\), \(\mpcstatename'\) such that \(\mpcstatename \stepstocon \mpcstatename'\),
and component \(\component\), if \(\pi_\context(\mpcstatename) = \domMap,\_\) and
\(\domMap ~ \component = \sealed{\_}\), then \(\pi_\mach(\mpcstatename) ~ \component =
\pi_\mach(\mpcstatename') ~ \component\).

@@

\definition Machine states \(\mach\) and \(\nach\) are {\em \(\components\)-variants} if, for
all \(\component \not \in \components\), \(\mach ~ \component = \nach ~ \component\). We write
this \(\mach \approx_\components \nach\).

\definition Machine state pairs \(\mach,\mach'\) and \(\nach,\nach'\) are said to change
together if for all components \(\component\) such that
\(\mach ~ \component \not = \mach' ~ \component\) or
\(\nach ~ \component \not = \nach' ~ \component\),
\(\mach' ~ \component = \nach' ~ \component\).
This is written \(\mach,\mach' \diamond \nach,\nach'\).

\definition A pair of traces \(\machT\) and \(\nachT\) are {\em in lockstep}
if their adjacent pairs change together. We will represent this relation with
\(\machT \doteq \nachT\). \bcp{People may not be familiar with the notation}
\sna{It's a new notation, we use this enough that it seems worth picking an
  evocative symbol.}\bcp{Sorry -- I meant the notational convention of representing a
coinductive relation with inference rules in this way, not the symbol...}
We define it as a coinductive relation:

@@

\definition Let \(\depth\) be any depth and \(f_\depth ~ \mach ~ (\_, \rts)\) hold for
\(|\rts| \geq \depth\). For any reachable segment \(\MPCT \langle f_\depth \rangle\),
let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
\(\components\) be the set of components \(\component\) such that \(\domMap ~ \component = \outside\).
\(\MPCT\) corresponds to the trace of a function from its call to its return.
Then take any \(\nach\) such that \(\mach \approx_\components \nach\), and
\(\nach \hookrightarrow \NPCT\langle f_\depth \rangle\). A system enjoys
{\em stepwise stack confidentiality} if for all such \(\MPCT\) and \(\NPCT\),
\(\MPCT \doteq \NPCT\).

@@

Example~\ref{fig:passingsimple} shows a sample trace of
a program that passes a value on the stack, without an attack.
We extend our call annotation with an additional parameter identifying a memory location to be
passed, and our diagram displays it in magenta. Our notion of domains is extended as well,
to include \(\passed{\depth,\depth'}\), recording two depths. We note the
current depth of the active function next to the stack pointer in the diagram, for reference.
The argument {\tt a} has
its value copied from {\tt x}, and we note that {\tt f} can read it -- because it is marked
as passed between depths 0 and 1, it is not varied. Then at instruction 13 it is even
modified, and naturally this is permitted as well.

@@

Note that in this policy both the entry and exit sequences
must write to each element of
the frame, which can be quite expensive, especially for programs that
allocate large but sparsely populated frames; we return to this point
in \cref{sec:lazy}.
%
For example, the program in \cref{fig:passingsimple} manages the stack
in this way. Here, the call pseudo-instruction to {\tt f} in line 5
hides a blessed sequence of instructions that allocate space to save
its return address to {\tt main} and seal the protected part of the
stack before {\tt f}'s body starts executing at 10. Conversely, the
return instruction in line 15 restores and jumps to the return address
and deallocates its frame space, and unseals the corresponding parts
of the stack. The function itself allocates and deallocates its own
private frame space in instructions 10 and 14. Additionally, unlike in
the simplified presentation of the policy, the caller shares the
arguments passed to the callee through the stack, and for this creates
and initializes (3--4) and destructs (6) the argument part of the
callee's stack frame.
%
%\rb{Note that we have also considered and implemented more flexible schemes
%  based on allowing operations above the current activation depth. How do we put
%  these together?}
%\rb{Do we want to discuss variations involving frame pointers or more implicit
%  return addresses, argument passing on the stack, alloc/dealloc, etc.?}
%\rb{Where to talk about the role of the compiler?}

>>>>

In the present work, we also distinguish between protecting stack data (like
LSE) and
control flow (with WBCF).  But rather than defining stack safety as ``what
a certain idealized machine does,'' we formalize explicit criteria
based on the technical framework of language-based
security\ifaftersubmission~\citep{??}\bcp{Maybe the Sabelfeld and Myers
  survey?}\fi.
Our core properties are \emph{stepwise stack integrity} and
\emph{stepwise stack confidentiality}. WBCF is a desireable property, but
we can express safety in an information-flow sense without reference to it,
even safety from attacks that hijack control flow.  \bcp{On consideration
  (especially in light of the weird ``C confuses A by returning directly to
  A instead of returning to B so that B can finish up some stuff before
  returning to A'' example\footnote{Actually, this example is not so weird.
  If C, the attacker, can return a {\em result} directly to A, who is
  expecting a result from B, then that is clearly bad!}), I
agree now that we should discuss WBCF.  It should be kept as distinct as
possible from the ``data safety'' part, since it is not really an
information flow property.  But our treatment of it is interestingly
different from StkTokens, because of our call annotations, and we should
discuss that.}

@@

We are finally ready for our definition of stack safety. We begin
by describing how we model knowledge about the program structure,
then dive into the formal definitions of local state encapsulation and
well-bracketed control flow, which together constitute stack safety\bcp{not
  any more?}.
Initially we will focus on a simple stack model, with a single stack
that doesn't share between caller and callee. All arguments are passed
in registers. Recall that we keep track of which addresses are
annotated as calls, with a ``code map'':
\[\codemap \in \CODEMAPS \subseteq \WORDS\]
\apt{Doesn't match previous definition any more.}

We parameterize over the structure of the stack, particularly the means by
which a caller identifies the data that should be protected during a call. We term
this the {\em sealing convention}, \(\sealcon \subseteq \MACHS \times \COMPONENTS\), which
maps a machine state to a set of addresses. \apt{Is it a relation or a function (seems to
  be the latter)?}
If a state's \(\PCname\) is annotated as a call, the sealing convention tells us
which addresses must be protected until the return. Similarly, we abstract over what
it means to return with a {\em return convention}, \(\retcon \subseteq \MACHS \times \MACHS\),
which relates call states to the states that can be considered valid returns to that call.
The usual return convention is that \(\mach\) and \(\mach'\) agree on the stack pointer,
and \(\mach'\) has its program counter advanced by one instruction from \(\mach\).
But whatever the return convention, at each step we update the context according
to the context-step function, \(\stepsto_C\).\apt{abrupt transition}

Each system component is assigned to a {\em domain}, which may be {\em Outside}
of the stack entirely, {\em Unsealed} and therefore accessible, or {\em Sealed}
at a particular depth. Each caller registers a {\em target}
condition -- a predicate on machine states that must hold for it to consider itself active.
A context is a pair of a ``domain map'' from components to domains, and a
``return target list,'' which is a stack of targets, one for each caller awaiting
return.  \ifaftersubmission\bcp{spacing is funny}\fi
%
\[\stackDom \in \STKDOMS ::= \outside + \unsealed + \sealed{\depth} \hspace*{0.5in} \target \subseteq \MACHS\]
\[(\domMap,\rts) = \context \in \CONTEXTS ::= (\COMPONENTS \rightarrow \STKDOMS)
  \times (\listT (\target)) \]
%
The initial context \(\context_0 = \domMap_0, []\), where \(\domMap_0\) maps
the stack to \(\unsealed\) and all other addresses and all registers to \(\outside\).

@@

When we build machine traces with this step function, we always know how deep in the
call stack we are and which components are sealed. We now proceed to state our properties.
The first is {\em stepwise stack integrity}, where ``stepwise'' means that the property is 
quantified over each step of execution from any initial state. The criterion for integrity
is simple: sealed components do not change.

\definition A system enjoys {\em stepwise stack integrity} if, for any reachable state
\(\mpcstatename\) such that \(\mpcstatename \stepstocon \mpcstatename'\),
and any component \(\component\), if \(\pi_\context(\mpcstatename) = \domMap,\_\) and
\(\domMap ~ \component = \sealed{\_}\), then \(\pi_\mach(\mpcstatename') ~ \component =
\pi_\mach(\mpcstatename) ~ \component\).

@@

\definition Machine states \(\mach\) and \(\nach\) are {\em \(\components\)-variants},
written \(\mach \approx_\components \nach\), if, for
all \(\component \not \in \components\), \(\mach ~ \component = \nach ~ \component\).

\definition Machine state pairs \(\mach,\mach'\) and \(\nach,\nach'\) are said to
\emph{change together}, written \(\mach,\mach' \diamond \nach,\nach'\), if,
for all components \(\component\) such that
\(\mach ~ \component \not = \mach' ~ \component\) or
\(\nach ~ \component \not = \nach' ~ \component\),
\(\mach' ~ \component = \nach' ~ \component\).

\definition A pair of traces \(\machT\) and \(\nachT\) are {\em in lockstep}
if their adjacent pairs of states change together. We will represent this relation with
\(\machT \doteq \nachT\). \bcp{People may not be familiar with the notation}
\sna{It's a new notation, we use this enough that it seems worth picking an
  evocative symbol.}\bcp{Sorry -- I meant the notational convention of representing a
coinductive relation with inference rules in this way, not the symbol...}
We define it as a coinductive relation:

@@

\definition For any depth \(\depth\), define \(f_\depth ~ \mach ~ (\_, \rts)\) to be 
\(|\rts| \geq \depth\). Then each reachable segment \(\MPCT \langle f_\depth \rangle\) 
corresponds to a trace of a function from its call to its return.
Given such a segment \(S\), let \(\mach,\_,(\domMap,\_)\)  = \(\head(S)\) and let 
\(\components\) be the set of components \(\component\) such that \(\domMap ~ \component = \outside\).
A system enjoys \emph{stepwise stack confidentiality} if, for all \(\MPCT\langle f_\depth\rangle\)
with associated \(m\) and \(\components\) as above, and 
for all \(\nach\) such that \(\mach \approx_\components \nach\) and
 \(\nach \hookrightarrow \NPCT\langle f_\depth \rangle\), 
 \(\MPCT\langle f_\depth\rangle \doteq \NPCT\langle f_\depth\rangle \).\apt{I found the previous
   version hopelessly confusing, but I'm not at all sure that this one is right! SNA please check.}
%% \definition Let \(\depth\) be any depth and \(f_\depth ~ \mach ~ (\_, \rts)\) hold for
%% \(|\rts| \geq \depth\). For any reachable segment \(\MPCT \langle f_\depth \rangle\),
%% let \(\head(\MPCT) = \mach,\pol,(\domMap,\rts)\) and
%%  \(\components\) be the set of components \(\component\) such that \(\domMap ~ \component = \outside\).
%% \(\MPCT\) corresponds to the trace of a function from its call to its return.
%% Then take any \(\nach\) such that \(\mach \approx_\components \nach\), and
%% \(\nach \hookrightarrow \NPCT\langle f_\depth \rangle\). A system enjoys
%% {\em stepwise stack confidentiality} if for all such \(\MPCT\) and \(\NPCT\),
%% \(\MPCT \doteq \NPCT\).

@@

Example \ref{fig:passingsimple} shows a sample trace of
a program that passes a value on the stack, without an attack.
We extend our call annotation with an additional parameter identifying a memory location to be
passed, and our diagram displays it in magenta. Our notion of domains is extended as well,
to include \(\passed{\depth,\depth'}\), recording two depths. We note the
current depth of the active function next to the stack pointer in the diagram, for reference.
The argument {\tt a} has
its value copied from {\tt x}, and we note that {\tt f} can read it -- because it is marked
as passed between depths 0 and 1, it is not varied. Then at instruction 13 it is even
modified, and naturally this is permitted as well.

@@

Note that in this policy both the entry and exit sequences
must write to each element of
the frame, which can be quite expensive, especially for programs that
allocate large but sparsely populated frames; we return to this point
in \cref{sec:lazy}.
%
The running example in \cref{fig:running-trace-a} manages the stack in this way.
For example, {\tt f}'s ``blessed'' sequences allocate space to save its return
address (and private variables) in instructions 20--21, and later restores that
return address and deallocates its private frame space before returning in
34--36. Additionally, unlike in the simplified presentation of the policy, the
caller passes information to the callee through the stack, and for this creates
and initializes (22 and 24--25) and destructs (29) the argument part of the
callee's stack frame.\rb{The policy would also need to control this, I think it
  is clear}
\rb{TODO: References to the running example are no longer valid! We could use
figures 5 or 6, referencing the figures, not the invisible assembly code.}
%
%\rb{Note that we have also considered and implemented more flexible schemes
%  based on allowing operations above the current activation depth. How do we put
%  these together?}
%\rb{Do we want to discuss variations involving frame pointers or more implicit
%  return addresses, argument passing on the stack, alloc/dealloc, etc.?}
%\rb{Where to talk about the role of the compiler?}
